Glyph-maps for Visually Exploring Temporal Patterns in Climate Data and Models

### 2012-03-07 revision/paper.tex
made data plural everywhere

Climate data is composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location (see many examples in \citet{IPCC}, like Figure 2.11). When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001}, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature (residuals from a linear model containing year and month terms), where red is mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \url{http://www.esrl.noaa.gov/psd/enso/enso.description.html}.) More work is required to see more localized patterns, like cooler land temperatures seen in early years.

@@

Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph of de-seasonalized temperature. The darker grey in the background represents the land mass of the study region. Because the data is de-seasonalized, the primary visible structure is still El Ni\~no, recognized by the bump in the middle of the series of the equatorial Pacific region. It is both a temporal anomaly and a spatial phenomenon, so this is not surprising. Other temporal features are visible from the many time series. Temperatures are generally more varied over land, than sea. A few locations, near the coast of South America and the southern part of North America, have dramatic jumps in temperature in the later years.%, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.

@@

\item[EXPO] The ASA 2009 data expo data \citep{murrell:2010} consists
  of monthly observations of several atmospheric variables from the
  International Satellite Cloud Climatology Project. The dataset
  includes observations over 72 months (1995--2000) on a 24 x 24 grid
  (576 locations) stretching from $113.75^{\circ}$W to
  $56.25^{\circ}$W longitude and $21.25^{\circ}$S to $36.25{^\circ}$N
  latitude.

@@

\end{itemize}
\end{quote}
Supplementary material contains the datasets and code to reproduce the plots in this paper.  R \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} was used.

@@

Each glyph is small, so there is not enough space for a full set of axes on each, and it is not desirable beause it would be repetitive, redundant information. Instead, minimal reference lines and boxes are incorporated. These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the data. The reference grid is a box, framing each icon, which represents the spatial grid. The reference line is a horizontal line at mid-range. Both help to read differences in slope and intercept. Slope is read by comparing the position of left and right endpoints on the frame, or by reading the angle between the data line and the reference line. Intercept is read from ``average'' position of the line in the box: is it near the top, or near the bottom? Figure~\ref{fig:ref-basic} demonstrates both guides. Reference elements are drawn in white, so they are minimally perceptible relative to the black data. Average temperatures were calculated for each month, and these were passed to the glyph calculations to produce the glyph-maps.   For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle (see Figure~\ref{fig:templates}).

@@

Climate data is usually large: many locations in space, and many points in time, which can make it arduous to work with computationally. The glyph-map calculations are as efficient as possible computationally, being linear in the number of spatial locations. It is possible that the calculations could also be parallelized for massive amounts of climate model data. Visualizing large data also hits a perceptual boundary.  There is a limit to the resolution of human visual system \citep{Kr12} which will prevent perfect perception of large spatial domains or seasonality of many years of data. Exploring large data requires reducing it to a resolution that can be reasonably absorbed visually.

Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP dataset, which is more realistically sized than the EXPO data. Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59. The most important feature to notice is the pattern of missing values. Most glyphs do not complete the full circle, typically missing points between 12 and 1, suggesting missing values in the earliest part of the data. Another important feature is variability, shown by the thickness of the circle. Some locations, particularly in the north and western mountain region, have large variations in temperature.

>>>>

Climate data are composed of multiple variables, such as temperature, ozone, precipitation, and winds, with a spatial and temporal context. The classic display for data of this type is a ``heatmap'', a tiled plot using color to display the value of the variable of interest at each location (see many examples in \citet{IPCC}, like Figure 2.11). When measurements are made at multiple time points it is reasonably common to display the data with small multiples~\citep{tufte:2001}, such as in Figure~\ref{fig:nasa-facet}. In this plot, a separate map is drawn for each month (columns) and year (rows), over a six year period of remotely sensed temperature data above Central America \citep{murrell:2010}. Color is used to display de-seasonalized temperature (residuals from a linear model containing year and month terms), where red is mapped to high values and blue to low. The most noticeable feature is the strong red patch in the equatorial Pacific beginning mid-1997 and tapering out during 1998. This is the El Ni\~no event, a major temperature anomaly (see for example, \citet{trenberth} or \url{http://www.esrl.noaa.gov/psd/enso/enso.description.html}.) More work is required to see more localized patterns, like cooler land temperatures seen in early years.

@@

Figure~\ref{fig:nasa-glyph} shows the glyph-map equivalent of Figure~\ref{fig:nasa-facet}: each location is represented by a small time series glyph of de-seasonalized temperature. The darker grey in the background represents the land mass of the study region. Because the data are de-seasonalized, the primary visible structure is still El Ni\~no, recognized by the bump in the middle of the series of the equatorial Pacific region. It is both a temporal anomaly and a spatial phenomenon, so this is not surprising. Other temporal features are visible from the many time series. Temperatures are generally more varied over land, than sea. A few locations, near the coast of South America and the southern part of North America, have dramatic jumps in temperature in the later years.%, all in high elevation areas, along the Andes and the Sierra Madre Occidental in Mexico.

@@

\item[EXPO] The ASA 2009 expo data \citep{murrell:2010} consists
  of monthly observations of several atmospheric variables from the
  International Satellite Cloud Climatology Project. The data set
  includes observations over 72 months (1995--2000) on a 24 x 24 grid
  (576 locations) stretching from $113.75^{\circ}$W to
  $56.25^{\circ}$W longitude and $21.25^{\circ}$S to $36.25{^\circ}$N
  latitude.

@@

\end{itemize}
\end{quote}
Supplementary material contains the data sets and code to reproduce the plots in this paper.  R \citep{R} with the packages {\tt ggplot2} \citep{me:ggplot2} and {\tt plyr} \citep{me:plyr} was used.

@@

Each glyph is small, so there is not enough space for a full set of axes on each, and it is not desirable beause it would be repetitive, redundant information. Instead, minimal reference lines and boxes are incorporated. These need to be minimally perceptible, post-attentive \citep{healey} and de-emphasized, so as not to detract from the data. The reference grid is a box, framing each icon, which represents the spatial grid. The reference line is a horizontal line at mid-range. Both help to read differences in slope and intercept. Slope is read by comparing the position of left and right endpoints on the frame, or by reading the angle between the data line and the reference line. Intercept is read from ``average'' position of the line in the box: is it near the top, or near the bottom? Figure~\ref{fig:ref-basic} demonstrates both guides. Reference elements are drawn in white, so they are minimally perceptible relative to the black data elements. Average temperatures were calculated for each month, and these were passed to the glyph calculations to produce the glyph-maps.   For star glyphs the reference frames are also useful and the equivalent of the reference line is a circle (see Figure~\ref{fig:templates}).

@@

Climate data are usually large: many locations in space, and many points in time, which can make it arduous to work with computationally. The glyph-map calculations are as efficient as possible computationally, being linear in the number of spatial locations. It is possible that the calculations could also be parallelized for massive amounts of climate model data. Visualizing large data also hits a perceptual boundary.  There is a limit to the resolution of human visual system \citep{Kr12} which will prevent perfect perception of large spatial domains or seasonality of many years of data. Exploring large data requires reducing it to a resolution that can be reasonably absorbed visually.

Figure~\ref{fig:gistemp-pred} shows the glyph-map for the GISTEMP data set, which is more realistically sized than the EXPO data. Glyphs at each location represent temperature from 1880--2011, starting with Jan 1880 at 12 o'clock and proceeding clockwise to 2011 at 11:59. The most important feature to notice is the pattern of missing values. Most glyphs do not complete the full circle, typically missing points between 12 and 1, suggesting missing values in the earliest part of the data. Another important feature is variability, shown by the thickness of the circle. Some locations, particularly in the north and western mountain region, have large variations in temperature.
